---
title: "Lotka-Volterra Example"
subtitle: "Numerical approximations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
library(kableExtra)
```

Day 3 morning session

Anna Ramji's version EDS212 Day 3 Computer Session 1

instructions: <https://allisonhorst.github.io/EDS_212_essential-math/comp_session_keys/comp_session_3a.html>

## **Fork and clone a repo**

In this session, you will use existing code in a public repo on GitHub that approximates the solution to the Lotka-Volterra equations.

You will make your own copy of the repo by *forking* it. By making a fork, you create a copy that you can work in separately.

-   Visit: [**https://github.com/allisonhorst/lotka-volterra-example**](https://github.com/allisonhorst/lotka-volterra-example)

-   Press the 'Fork' button to make a copy - note the repo is now copied into your account

-   So far, you just have a remote copy. We need a local copy of the repo, that we can work in locally. We do that by *cloning* a repo. Press the 'Code' button in YOUR copy of the repo. Copy the URL.

-   Go back to RStudio. Create a new project WITH version control. Paste the URL where prompted, then choose the destination for your new project. Create the project.

-   You should see that the materials from the remote repo now exist locally in a version controlled R project. Yay!

## **Run existing code**

The repo contains a single .Rmd (lotka-volterra.Rmd), with information and code to numerically approximate solutions to the Lotka-Volterra equations and make a graph.\
\
Open the .Rmd and knit. We will talk through the code together. A couple of things to note:

-   When you knit, the .html doesn't appear in the git pane. Why?

-   How does the numerical approximation change if you update the initial conditions? The parameter estimates?

-   Try making changes, then stage \> commit \> push back to your remote repo. Make sure the changes appear there.

-   Congratulations! You've adding *forking* and *cloning* to your git toolkit.

## **Command line intro + git**

Here, we'll learn a few basic commands to talk with your computer through the command line to navigate around and do some git stuff outside of RStudio.

Some basics:

-   Open the Terminal (outside of RStudio)

-   Use pwd to see where your Terminal currently sees the working directory

-   Use ls to see the contents of the directory

-   Use cd to change to a different downstream directory

-   Use .. to go back upstream

Now for git:

-   Navigate to your current project directory

-   Use git status to see the current git status

-   Make a change to the .Rmd in your project and save

-   Use git add filename.Rmd to stage it (git add . will stage everything)

-   Use git commit -m "a commit message" to commit to local repo

-   Use git checkout again to check status

-   Use git pull to pull

-   Use git push to push changes

-   Check with git checkout

## **Below: additional material ([not]{.underline} done in class)**

In this session, we'll see how to use the `deSolve::ode()` function to numerically estimate the solution to the Lotke-Volterra equations for predator-prey populations.

First, we'll learn a bit more about how numerical methods for approximating solutions to differential equations work.

**Credit:** This lesson is closely based on the article (Numerically solving differential equations with R)[[**https://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html**](https://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html)]

```         
library(tidyverse) library(deSolve) library(kableExtra)
```

## Content Below \~\~\~

**Credit:** This code is closely based on the article [Numerically solving differential equations with R](https://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html)

### The Lotka-Volterra equations

As described in the lecture, the Lotke-Volterra models have been used to describe predator-prey populations.

#### Prey equation:

$$\frac{dx}{dt}=\alpha x-\beta xy$$

From Wikipedia: "The prey are assumed to have an unlimited food supply and to reproduce exponentially, unless subject to predation; this exponential growth is represented in the equation above by the term $\alpha x$. The rate of predation upon the prey is assumed to be proportional to the rate at which the predators and the prey meet, this is represented above by $\beta xy$. If either x or y is zero, then there can be no predation."

#### Predator equation:

$$\frac{dy}{dt}=\delta xy - \gamma y$$

From Wikipedia: "In this equation, $\delta xy$ represents the growth of the predator population. (Note the similarity to the predation rate; however, a different constant is used, as the rate at which the predator population grows is not necessarily equal to the rate at which it consumes the prey). The term $\gamma y$ represents the loss rate of the predators due to either natural death or emigration, it leads to an exponential decay in the absence of prey.

Where:

-   $x$ is prey population (e.g. rabbits)
-   $y$ is predator population (e.g. wolves)
-   $\alpha, \beta, \gamma, \delta$ are positive parameters

To find an approximate solution in R, we will need four things: - Parameter values - A sequence of times over which we'll approximate predator & prey populations - An initial condition (initial populations of predator & prey at t = 0) - The differential equations that need to be solved

Solving the Lotke-Volterra equation:

```{r}
# Create a sequence of times (days): 
time <- seq(0, 25, by = 0.05)

# Set some parameter values (these can change - keep it in mind):
parameters <- c(alpha = .75, beta = 0.8, delta = 0.5, gamma = 1)

# Set the initial condition (prey and predator populations at t = 0).
# Recall: x = prey, y = predator
init_cond <- c(x = 5, y = 5)

# Prepare the series of differential equations as a function: 
lk_equations <- function(time, init_cond, parameters) {
  with(as.list(c(init_cond, parameters)), {
    dxdt = alpha * x - beta * x * y
    dydt = delta * x * y - gamma * y
    return(list(c(dxdt, dydt)))
  })
}

# Find the approximate the solution using `deSolve::ode()`:
approx_lk <- ode(y = init_cond, times = time, func = lk_equations, parms = parameters)

# Check the class: 
class(approx_lk)

# We really want this to be a data frame, and we want both prey (x) and predator (y) to be in the same column -- we'll learn why in EDS 221 (tidy data)

# pivot_longer() function used to turn approx_lk from having x (rabbits) and y (wolves) as separate columns showing numeric values into having a combined "species" column with x and y and their values into a new column called "population" 

# * . html --> * is an inclusive operator telling git ignore to ignore everything and anything .html


approx_lk_df <- data.frame(approx_lk) %>% 
  pivot_longer(cols = c(x,y), names_to = "species", values_to = "population")

# Plot it! 
ggplot(data = approx_lk_df, aes(x = time, y = population)) +
  geom_line(aes(color = species))
```

## Updating parameter values

What happens as you change the different parameters (and re-run the entire code chunk)? How does that align with what you see in the graph? Some things to keep in mind:

-   $\alpha$ is a growth rate for prey
-   $\gamma$ is a mortality rate for predator

## End
